\paperTitle

### 2014-05-16 doc/icml2014/undirected.tex

Of course, the log-likelihood is non-concave:
\begin{align}
L_\text{unsup}(\theta) \eqdef \E_{\vx \sim \sD}[\log \sum_{\vh \in \sH} p_\theta(\vx,\vh)].
\end{align}
% PL: this doesn't make sense
%While $\phi_\sC(x,h)$ can be an arbitrary featurization, we assume
%  without loss of generality that $\phi_\sC(x,h)$ are the indicator
%  features $\BI[x,h]$\footnote{Note that $\BI[x,h]$ are sufficient
%  statistics for the model, and hence we do not lose any statistical
%  efficiency in using them.}.

%In this section, we will show that we can extend the results to
  %undirected graphical models parameterized as log-linear latent variable
  %models.
On the other hand, 
if we were able to observe $\vh$, the \emph{supervised} likelihood would be concave:
\begin{align}
\label{eqn:logLinearSupervised}
L_\text{sup}(\theta) &\eqdef \E_{(\vx,\vh) \sim \sD_\text{sup}}[\log p_\theta(\vx,\vh)] \\
                     &= \theta^\top \left(\sum_{\sC \in \sG} \E[\phi(\vx_\sC,\vh_\sC)]\right) - A(\theta).
\end{align}
While we don't have supervised data,
note that the method of moments gives us hidden marginals $Z_\sC = \BP(\vh_\sC)$.
Furthermore, given $\mOpp{v}{i}$ and $Z_\sC$ of any $\sC \ni v$, we can recover
all marginals $\BP(\vx_\sC, \vh_\sC) = Z_\sC \cdot \mOppAll[\vx_\sV]$.%\todo{clean up}.
Finally, the key step is that we can compute all the expected feature vectors from the marginals:
\begin{align}
\label{eqn:logLinearFeatures}
\mu_\sC \eqdef \E[\phi(\vx_\sC,\vh_\sC)] = \sum_{\vx_\sC,\vh_\sC} \BP(\vx_\sC,\vh_\sC) \phi(\vx_\sC,\vh_\sC).
\end{align}
Therefore, we can optimize the supervised likelihood objective without actually
having supervised data.
Of course, in the finite data regime, the method of moments yields estimates
$\hat \mu^\text{mom}_\sC$ which approximate the true $\mu_\sC$.
In supervised learning, we have different estimate $\hat\mu^\text{sup}_\sC$ of $\mu_\sC$ based on an empirical average
over the data points.
In the limit of infinite data, both estimators converge to $\mu_\sC$.

@@

%However, given the moments $\E[ \phi_\sC(\vx,\vh) ] = Z_\sC$, the
%  problem simplifies to,
%\begin{align}
%  \hat \theta 
%  &= \arg\max_{\theta} \sum_{\sC \in \sG} \theta_\sC^\top Z_\sC - A(\theta), \label{eqn:obj}
%\end{align}
%which is concave in $\theta$.
\algorithmref{undirected} summarizes our approach.
In \sectionref{directed}, we showed that $\LearnMarginals$ is a consistent
estimator for $Z_\sC$ for any graphical model satisfying the exclusive views property (\propertyref{exclusive-views}). 
Consequently, the result is that the output of \algorithmref{undirected},
$\hat\theta$, is a consistent estimator.\footnote{Of course, the log-linear model must be identifiable
(it is violated, for example, if two components of $\phi(\vx_\sC, \vh_\sC)$ were identical).
In general, we can identify the parameters up to the kernel of $\nabla^2 A(\theta)$.}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnLogLinear}
  \label{algo:undirected}
  \begin{algorithmic}
    \REQUIRE A graphical model $\sG$ satisfying \propertyref{bottleneck}, data $\sD$
    \ENSURE A consistent estimate of parameters $\hat \theta$
      \STATE Apply $\LearnMarginals$ to $\sG$ to learn clique marginals $Z_\sC$.
      \STATE Compute expected features (\equationref{logLinearFeatures}) and optimize \equationref{logLinearSupervised}.
  \end{algorithmic}
\end{algorithm}

\paragraph{Pseudolikelihood.}
Unlike for directed models, in the undirected case,
graphs with high treewidth present a challenge
for even fully-supervised maximum likelihood due to the intractability of the log-partition function.
This motivates looking at pseudolikelihood objective of \citep{besag75pseudo} for a computationally more
efficient solution.
The pseudolikelihood objective sums over the log-probability of each variable $a$ given its neighbors $\sN(a)$:
$$L_\text{pseudo}(\theta) \eqdef \E_{x \sim \sD}\left[\sum_{a \in X \cup H} \log p_\theta(a \mid x_{\sN(a)})\right].$$
In the fully-supervised setting, it is well-known that pseudolikelihood provides computationally efficient consistent estimates,
albeit at the cost of statistical efficiency.
Coincidentally, this is the same high-level motivation for using method of moments in the first place.

The pseudolikelihood also replaces the log-partition function
  $A(\theta)$ with a sum over the variables in the neighborhood:
  $A_a(\theta; x_{\sN(a)}) = \sum_{a' \in \{a\} \union \sN(a)}
  \exp(\theta^\top\phi_{\sN(a)}(a'))$.
Note that the neighbors $\sN(a)$ in general do not form a clique, and
  thus the uniformly bottlenecked assumption alone is not sufficient to
  guarantee exclusive views for each variable in the subgraph.
However, when we do have exclusive views for each neighborhood, as in
  \figureref{examples-mrf}, we can adapt our earlier algorithm
  ($\LearnClique$) to estimate the marginals over all neighborhoods $Z_{\sN(a)}$.

>>>>

Of course, the log-likelihood is non-concave:
\begin{align}
  L_\text{unsup}(\theta) \eqdef \E_{\vx \sim p^*}[\log \sum_{\vh \in \sH} p_\theta(\vx,\vh)],
\end{align}
where we are assuming for simplicity that we have infinite data drawn from the true distribution $p^*$.
% PL: this doesn't make sense
%While $\phi_\sC(x,h)$ can be an arbitrary featurization, we assume
%  without loss of generality that $\phi_\sC(x,h)$ are the indicator
%  features $\BI[x,h]$\footnote{Note that $\BI[x,h]$ are sufficient
%  statistics for the model, and hence we do not lose any statistical
%  efficiency in using them.}.

%In this section, we will show that we can extend the results to
  %undirected graphical models parameterized as log-linear latent variable
  %models.
On the other hand, 
if we were able to observe $\vh$, we would optimize \emph{supervised} likelihood, which is concave:
\begin{align}
\label{eqn:logLinearSupervised}
L_\text{sup}(\theta) &\eqdef \E_{(\vx,\vh) \sim p^*}[\log p_\theta(\vx,\vh)] \\
                     &= \theta^\top \left(\sum_{\sC \in \sG} \E[\phi(\vx_\sC,\vh_\sC)]\right) - A(\theta).
\end{align}
While we don't have supervised data,
recall that the method of moments provides the hidden marginals $Z_\sC = \BP(\vh_\sC)$
and conditional moments $\mOpp{v}{i} = \Pr(x_v \mid h_i)$.
Assuming our graphical model is in canonical form (\lemmaref{reduction}),
every observed variable is connected to exactly one hidden variable.
Then $Z_\sC$ and $\mOpp{v}{i}$ directly yield marginals for every clique $\BP(\vx_\sC, \vh_\sC)$.
Given these marginals, we can compute the expected local feature vector:
\begin{align}
\label{eqn:logLinearFeatures}
\mu_\sC \eqdef \E[\phi(\vx_\sC,\vh_\sC)] = \sum_{\vx_\sC,\vh_\sC} \BP(\vx_\sC,\vh_\sC) \phi(\vx_\sC,\vh_\sC).
\end{align}
Note that $\{\mu_\sC\}_{\sC \in \sG}$ are exactly the sufficient statistics of the model (\equationref{logLinearSupervised}),
so we can optimize the supervised likelihood objective without actually having any supervised data!
In the finite data regime, the method of moments yields the estimate
$\hat \mu^\text{mom}_\sC$ which approximates the true $\mu_\sC$.
In supervised learning, we obtain a different estimate $\hat\mu^\text{sup}_\sC$ of $\mu_\sC$ based on an empirical average
over data points.
In the limit of infinite data, both estimators converge to $\mu_\sC$.

@@

%However, given the moments $\E[ \phi_\sC(\vx,\vh) ] = Z_\sC$, the
%  problem simplifies to,
%\begin{align}
%  \hat \theta 
%  &= \arg\max_{\theta} \sum_{\sC \in \sG} \theta_\sC^\top Z_\sC - A(\theta), \label{eqn:obj}
%\end{align}
%which is concave in $\theta$.
\algorithmref{undirected} summarizes our approach.
In \sectionref{directed}, we showed that $\LearnMarginals$ is a consistent
estimator for $Z_\sC$ for any graphical model satisfying the uniformly bottlenecked property (\propertyref{bottleneck}). 
Consequently, the output of \algorithmref{undirected}
is a consistent estimator.\footnote{Of course, the log-linear model must be identifiable
(it is violated, for example, if two components of $\phi(\vx_\sC, \vh_\sC)$ were identical).
In general, we can identify the parameters up to the kernel of $\nabla^2 A(\theta)$.}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnLogLinear}
  \label{algo:undirected}
  \begin{algorithmic}
    \REQUIRE A graphical model $\sG$ satisfying \propertyref{bottleneck}, data $\sD$
    \ENSURE Parameters $\theta$
      \STATE Apply $\LearnMarginals$ to $\sG$ to learn clique marginals $Z_\sC$.
      \STATE Compute expected features $\{\mu_\sC\}_{\sC \in \sG}$ (\equationref{logLinearFeatures}) and optimize \equationref{logLinearSupervised}.
  \end{algorithmic}
\end{algorithm}

\paragraph{Pseudolikelihood.}
Unlike for directed models, in the undirected case,
graphs with high treewidth present a challenge
for even fully-supervised maximum likelihood due to the intractability of the log-partition function $A(\theta)$.
%One can employ various variational approximations of $A(\theta)$ \cite{wainwright08varinf},
This motivates looking at pseudolikelihood objective of \citep{besag75pseudo} for a computationally more
efficient solution.
The pseudolikelihood objective is a sum over the log-probability of each variable $a$ given its neighbors $\sN(a)$:
\begin{align}
  \label{eqn:pseudolikelihood}
L_\text{pseudo}(\theta) \eqdef \E_{(\bx,\bh) \sim p^*}\left[\sum_{a \in X \cup H} \log p_\theta(a \mid \sN(a))\right].
\end{align}
In the fully-supervised setting, it is well-known that pseudolikelihood provides computationally efficient consistent estimates,
albeit at the cost of statistical efficiency.
Coincidentally, this is the same high-level motivation for using method of moments in the first place.

Let $\phi_{a,\sN(a)}(a, \sN(a)) = \sum_{\sC \ni a} \phi_\sC(\bx_\sC,\bh_\sC)$
denote the sum over cliques $\sC$ that contain $a$; note that only depends on $a$ and its neighbors $\sN(a)$.
We can write each conditional log-likelihood as:
\begin{align}
p_\theta(a \mid \sN(a)) = \exp(\theta^\top\phi_{a,\sN(a)}(a, \sN(a)) - A_a(\theta; \sN(a))),
\end{align}
where the conditional log-partition function $A_a(\theta; \sN(a)) = \log \sum_{a'} \exp(\theta^\top\phi_{a,\sN(a)}(a', \sN(a)))$
involves summing only over values to a single variable.

If we knew $\mu_{a,\sN(a)} \eqdef \E[\phi_{a,\sN(a)}(a,\sN(a))]$,
then we would be able to optimize the pseudolikelihood objective again without
having access to any labeled data.
%The pseudolikelihood also replaces the log-partition function
%  $A(\theta)$ with a sum over the variables in the neighborhood:
%  $A_a(\theta; x_{\sN(a)}) = \sum_{a' \in \{a\} \union \sN(a)}
%  \exp(\theta^\top\phi_{\sN(a)}(a'))$.
Note that $\{a\}\cup\sN(a)$ in general do not form a clique.
Fortunately, \lemmaref{bottleneck-views} holds for all biconnected subgraphs.
We can break $\{a\}\cup\sN(a)$ into a collection of 
  % PL: this is not necessary
  %thus the uniformly bottlenecked assumption alone is not sufficient to
  %guarantee exclusive views for each variable in that subgraph.
%However, when we do have exclusive views for each neighborhood, as in
  \figureref{examples-mrf},
  Therefore, we can adapt our algorithm ($\LearnClique$) to estimate all marginals of the form $Z_{a,\sN(a)} \eqdef \Pr(a,\sN(a))$,
  from which we can compute the features $\mu_{a,\sN(a)}$,
  and optimize \equationref{pseudolikelihood} to obtain a consistent estimate of $\theta$.
